The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-04 paper.tex
Pequenos ajustes + Figuras do Adriano

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

@@

    <...>  // other labels
  }
}
\end{lstlisting}
\end{minipage}
\caption{
%{\small %\textmd{
Static allocation of trails, entry-point labels, and dispatch loop.
%}%}
\label{lst.impl.trails}
}
\end{figure}

@@

#ifdef CEU_WCLOCKS
    event void Timer.fired () {
        int dt = 10000;
        ceu_sys_go(EVT__WCLOCK, &dt);
    }
#endif

#ifdef _EVT_PHOTO_READDONE
    event void Photo.readDone (uint16_t val) {
        ceu_sys_go(EVT_PHOTO_READDONE, &val);
    }
#endif

#ifdef _EVT_RADIO_SENDDONE
    event void RadioSend.sendDone (message_t* msg) {
        ceu_sys_go(EVT_RADIO_SENDDONE, &msg);
    }
#endif

#ifdef _EVT_RADIO_RECEIVE
    event message_t* RadioReceive.receive (message_t* msg) {
        ceu_sys_go(EVT_RADIO_RECEIVE, &msg);
        return msg;
    }
#endif

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
that triggers callback \code{Timer.fired} (lines 13--18) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 16)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 20--24) and the radio transceiver 
(lines 26--37).

@@

The communication between scripts in \CEU and the components in the \VM is 
mostly through events:
scripts \code{emit} requests through \code{output} events and \code{await} 
answers through \code{input} events.
%
Terra also provides system calls for initialization and configuration of 
components.

@@

>>>>

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input int A, B, C;
do
    var int a = await A;
end
do
    var int b = await B;
end
par/and do
    await B;
with
    await C;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
union {             // sequence
    int a_1;        //   do_1
    int b_2;        //   do_2
    struct {        //   par/and
        int _and_3: 1;
        int _and_4: 1;
    };
} MEM ;
\end{lstlisting}
\end{minipage}
%\rule{14cm}{0.37pt}
\caption{
A program with blocks in sequence and in parallel, with corresponding memory 
layout that the compiler generates.
{\small %\textmd{
}%}
\label{lst.impl.mem}
}
\end{figure}

@@

    <...>  // other labels
  }
}
\end{lstlisting}
\end{minipage}
\caption{
%{\small %\textmd{
From left to right: static allocation of trails, entry-point labels, and 
dispatch function.
In the left, the comments identify the trail indexes inferred by the compiler.
In the middle, each trail segment has an associated numeric identifier 
generated by the compiler.
In the right, the dispatcher uses a \code{switch} to associate each segment 
identifier with the corresponding code to execute.
%}%}
\label{lst.impl.trails}
}
\end{figure}

@@

#ifdef CEU_WCLOCKS
    event void Timer.fired () {
        ceu_sys_wclock(10000);
    }
#endif

#ifdef EVT_PHOTO_READDONE
    event void Photo.readDone (uint16_t val) {
        ceu_sys_go(EVT_PHOTO_READDONE, &val);
    }
#endif

#ifdef EVT_RADIO_SENDDONE
    event void RadioSend.sendDone (message_t* msg) {
        ceu_sys_go(EVT_RADIO_SENDDONE, &msg);
    }
#endif

#ifdef EVT_RADIO_RECEIVE
    event message_t* RadioReceive.receive (message_t* msg) {
        ceu_sys_go(EVT_RADIO_RECEIVE, &msg);
        return msg;
    }
#endif

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
%
The file \code{ceu.h} (included in line 3) contains all definitions for the 
compiled \CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (included in line 4) contains the main loop of \CEU 
pointing to the labels defined in the program.
The callback \code{Boot.booted} (lines 6--11) is called by TinyOS on startup, 
so we initialize \CEU inside it (line 7).
If the \CEU program uses timers, we also start a periodic timer (lines 8--10) 
that triggers callback \code{Timer.fired} (lines 13--17) every 10 milliseconds 
and advances the wall-clock time of \CEU (line 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (lines 19--23) and the radio transceiver 
(lines 25--36).

@@

The communication between scripts in \CEU and the components in the \VM is 
mostly through events:
scripts \code{emit} requests through \code{output} events and \code{await} 
answers through \code{input} events.
%
Terra also provides system calls for initialization and configuration of 
components.
%
TODO: explicar Figura 21.
\begin{comment}
Figure~\ref{lst.terra.defs} illustrates a program interface declaration in \CEU
(in the left), and associated binding code in \C

 with respective binding customization scenario.
This customization sample uses a small data structure, two output events, one input event, and two functions.
The left listing is the definition file to be included into the \CEU script program and the right listing is the
definition file to be included into the VM custom module implementation.
\end{comment}

\begin{figure}
\begin{minipage}[t]{0.46\linewidth}
\begin{lstlisting}[morekeywords={ubyte,ushort,output,function,regtype}]
// Message definition
regtype sensorMsg with
    var ubyte  id;
    var ushort value;
end

// Output events definition
output void REQ_TEMP void          1;
output void SEND_SENSOR sensorMsg  2;

// Input events definition
input void TEMP void               1;

// functions definitions
function ubyte getNodeId()         1;
function ubyte queuePut(sensorMsg) 2;
.
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.51\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
morekeywords={nx_struct,nx_uint8_t,nx_uint16_t,uint8_t,uint32_t}]

typedef nx_struct sensorMsg{
  nx_uint8_t  id;
  nx_uint16_t value;
} sensorMsg_t;

enum {
  O_REQ_TEMP    = 1;
  O_SEND_SENSOR = 2;

  I_TEMP        = 1;

  F_GETNODEID   = 1;
  F_QUEUEPUT    = 2;
};

\end{lstlisting}
\end{minipage}
\caption{
%{\small %\textmd{
Terra control block sample and the equivalent definition for VM implementation.
%}%}
\label{lst.terra.defs}
}
\end{figure}

TODO: dizer o que eh manual/automatico nos lsts de ceu

@@

\begin{comment}
%-------------------------------------------------
Figure~\ref{lst.impl.custom} presents a partial implementation for the elements 
introduced in the Figure~\ref{lst.impl.defs}.
The two first functions are called from the VM decoder, the fist function \code{procOutevt()} dispatch any defined output events
and the function \code{callFunction()} dispatch any defined customized functions.
Lines 17--22 shows an example of a custom function that returns a value via stack.
Lines 24--27 has a example of an external event call.
Lines 29--34 presents an example how a input event is queued to the VM engine.
\end{comment}

%-----------------------------------------------------
% Equivalent to the fig. 19

\begin{figure}
\begin{minipage}[t]{0.81\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=2.5em,
morekeywords={uint8_t,uint16_t,uint32_t,error_t,command}]
// Output event dispatcher
command void VM.procOutEvt(uint8_t id, uint32_t value){
  switch (id){
    case O_REQ_TEMP    : proc_req_temp(id,value);    break;
    case O_SEND_SENSOR : proc_send_sensor(id,value); break;
  }
}

// Function dispatcher
command void VM.callFunction(uint8_t id){
  switch (id){
    case F_GETNODEID : func_getNodeId(id); break;
    case F_QUEUEPUT  : func_queuePut(id);  break;
  }
}

// Pushing a value to the stack
void  func_getNodeId(uint16_t id){
  uint16_t stat;
  stat = TOS_NODE_ID;
  signal VM.push(stat);
}

// Calling a output event
void  proc_req_temp(uint16_t id, uint32_t value){
  call S_TEMP.read();
}

// Queueing an input event + value
uint16_t lastTemp;
event void S_TEMP.readDone(error_t result, uint16_t val)
  lastTemp = val;
  signal VM.queueEvt(I_TEMP, 0, &lastTemp);
}

\end{lstlisting}
\end{minipage}

\caption{
%{\small %\textmd{
VM Customization -- input/output events and functions.
%}%}
\label{lst.impl.custom}
}
\end{figure}

%------------------------------------------------------
